---
title: "Severe Weather Event Report"
author: "SN de Koning"
date: "21 februari 2016"
output: html_document
---

## Synopsis
In this report we will examine the effects of extreme weather effects across the United States. It will try to answer the following two questions, namely which weather events are the most harmful with regard to public health, and which events carry the most economic consequences.

## Data Processing

The storm database that was used was from the U.S. National Oceanic and Atmospheric Administration's (NOAA). and was aquired as follows:

```{r}
library(dplyr)
```

```{r, cache = TRUE}
getdata <- function(){
    if(!file.exists("storm_data.csv.bz2")){
        file_url <- paste("https://d396qusza40orc.cloudfront.net/",
                    "repdata%2Fdata%2FStormData.csv.bz2",
                    sep = "")
        download.file(file_url, "storm_data.csv.bz2")
    }
    else print("Data already present")
}

getdata()

stdata <- tbl_df(read.csv(bzfile("storm_data.csv.bz2")))
```

String and character imput contained capitalization differences, and were transformed to be all uppercase. Futher the summary events were dropped.
```{r}
stdata_corr <- stdata %>% mutate(EVTYPE = toupper(EVTYPE), 
                 PROPDMGEXP = toupper(PROPDMGEXP), 
                 CROPDMGEXP = toupper(CROPDMGEXP))

ind_sum <- grep("SUMMARY", stdata_corr$EVTYPE)
stdata_corr <- stdata_corr[-ind_sum,]

stdata_corr <- select(stdata_corr, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
                    CROPDMG, CROPDMGEXP)
```
From this a data table was constructed containing the variables with the type of weather event "EVTYPE", and the variables pertaining to public health "FATALITIES", and "INJURIES". Entries containing no fatalities or injuries were dropped.

```{r}
stdata_pubdmg <- stdata_corr %>% select(EVTYPE, FATALITIES, INJURIES) %>%
    filter(FATALITIES != 0 | INJURIES != 0)
```

The datafile contained two variables which served as multiplier for the financial damages. These multipliers were coded as a character denoting the level of multiplication, i.e. "H" = 1 x 10^2 (Hundreds), "K" = 1 x 10^3 (Thousands), "M" = 1 x 10^6 (Millions) and "B" = 1 x 10^9 (Billions).

However, these variables also contained a small amount of non-standard multipliers. As these were a negligeble proportion of the total amount of multipliers, these were considered input errors, and as such dropped from the construction of the financial damage data table, and as such from the analysis.

```{r}
nonstd <- nrow(subset(stdata_corr, PROPDMGEXP == "+" | PROPDMGEXP == "0" | PROPDMGEXP == "5"| PROPDMGEXP == "6" | PROPDMGEXP == "?" | PROPDMGEXP == "4" | PROPDMGEXP == "2" | PROPDMGEXP == "3" | PROPDMGEXP == "7" | PROPDMGEXP == "-" | PROPDMGEXP == "1" | PROPDMGEXP == "8" | CROPDMGEXP == "?" | CROPDMGEXP == "0" | CROPDMGEXP == "2"))

nonstd / nrow(stdata_corr)
```

The data table for financial damages was constructed as follows
```{r}
exp_transform <- function(damages, exponential){
    if(exponential == ""){
        damages * 1
    }
    else if(exponential == "K"){
        damages * 1000
    }
    else if(exponential == "M"){
        damages * 1E6
    }
    else if(exponential == "B"){
        damages * 1E9
    }
}

stdata_ecodmg <- stdata_corr %>% 
    select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
    filter(PROPDMG != 0 | CROPDMG != 0)
 
stdata_ecodmg <- stdata_ecodmg %>%
    filter(PROPDMGEXP == "" | PROPDMGEXP == "K" | PROPDMGEXP == "M" | PROPDMGEXP == "B") %>% 
    filter(CROPDMGEXP == "" | CROPDMGEXP == "K" | CROPDMGEXP == "M" | CROPDMGEXP == "B")

stdata_ecodmg <-mutate(stdata_ecodmg, 
                        PROPDMG_TOT = exp_transform(stdata_ecodmg$PROPDMG, stdata_ecodmg$PROPDMGEXP), 
                        CROPDMG_TOT = exp_transform(stdata_ecodmg$CROPDMG, stdata_ecodmg$CROPDMGEXP))
```

## Results
```{r}
library(ggplot2)
```

Which type of weather event costs the most by way of public health damages?

```{r}
## some code
```

Which type of weather event costs the most in financial damages?

```{r}
## Some code
```

## Appendix

*Sessioninfo and Date*
```{r}
software_environment <-sessionInfo()
print(software_environment)
Sys.Date()
```